FAIR fission track analysis with geochron@home

### 2025-08-21 geochron@home/manuscript/paper.tex
Changes to outlook section

\begingroup
\captionof{table}{ Summary table of the crowd sourcing results for the
  different users (rows) and grains (columns).  The last row is the
  ratio of the median count to PV's count (user~1) for each grain.
  The last column is the ratio of the weighted mean count over grains
  compared with PV for the same grains.
}
\label{tab:crowdtable}
\renewcommand{\arraystretch}{0.7}
\begin{scriptsize}
  \input{../output/crowdtable.txt}
\end{scriptsize}
\endgroup

In 1906, Sir Francis Galton visited a county fair in which a contest
was held to guess the weight of an ox. 787 people participated in the
event. Galton discovered that the median of all their estimates was
within 0.8\% of the true weight of the ox and more accurate than 90\%
of the individual estimates. Such is the `wisdom of crowds'
\citep{galton1907a}. Similar effects are seen in fission track
geochronology. An interlaboratory comparison study by
\citet{miller1985} showed that the average of several fission track
age estimates is closer to the known age of mineral standards than the
age obtained by any individual observer. Inspired by these previous
experiments, we used geochron@home to bring fission track analysis to
a proverbial `county fair' of citizen scientists.\medskip

We here present the results of a type of `crowd-sourcing' experiment
that was carried out at UCL as part of an undergraduate course in
isotope geology. 70 students were asked to create an account on the
geochron@home server by selecting a unique username and
password. After spending a few minutes to complete the compulsory
tutorial (Section~\ref{sec:tutorial}), the students were asked to
count fission tracks in randomly assigned image stacks of Mount
Dromedary apatite. Each student was required to analyse at least 15
grains, using pre-defined ROIs. In a matter of hours, the students
amassed a dataset of 37,331 fission track counts in 34 grains.  This
large dataset was parsed into separate data files --- one for each
student --- which they processed during an assessed programming
exercise.\medskip

@@

However, in spite of the under-counting by the students, the
\emph{ratios} of the track counts between different grains are
generally quite close to PV's ratios, especially when comparing grains
with similar track densities. This is apparent from
Table~\ref{tab:crowdtable} (looking along columns).
Figure~\ref{fig:23vs25}b shows a scatter plot of pairs of counts for
two grains (23 and 25) made by 36 students who counted both of
them. They average around 29 and 30 tracks per grain compared with
PV's pair of 52 and 50, which is highlighted in blue. There is a
positive correlation between the pairs, consistent with systematic
observer effects (i.e., people who count a low value in one grain tend
to count similarly low in the other, and vice versa).  Comparing other
pairs of grains shows similar results, with correlations varying
between 0.4 and 0.9. In all cases we have looked at, the residual
variation, after allowing for systematic differences between students,
is consistent with Poisson variation.

As mentioned in Section~\ref{sec:architecture}, geochron@home stores
the actual track positions marked by the users. This raw data can be
downloaded as a \texttt{.json} file and inspected in detail.
Figure~\ref{fig:23vs25}b) shows two-dimensional histograms for the x-y
positions of all the track positions generated by the students in the
two apatite crystals. Visual comparison of the histogram with the
optical image confirms that the students unanimously identified the
most obvious semitracks, which contain a clearly visible etch pit and
tail. Shorter and fainter tracks received fewer clicks. Datasets like
this can be used to replace integer counts of fission tracks with
probabilities, reflecting the ambiguity of some fission track
datasets. Figure~\ref{fig:23vs25} also shows that some students
counted the tails of the fission tracks rather than their etch pits,
despite being told the opposite in the tutorial. Fixing this issue
will require some improvements to the tutorial pages
(Section~\ref{sec:outlook}).\medskip

%\begin{figure}[!ht]
{ \centering \includegraphics[width=9cm]{23vs25.pdf}
  \captionof{figure}{ a) The track counts for grain~25 plotted against
    those for grain~23 for 37 students who counted both of those
    grains. b) Optical images in transmitted light of Mount Dromedary
    apatites 23 and 25 in the crowd-sourcing experiment.  c,d)
    Two-dimensional histogram of the track locations for the two
    grains, as identified by the citizen scientists.
    \medskip}
  \label{fig:23vs25}
}%\end{figure}

@@

\begin{enumerate}
\item Interactive tutorial pages. To ensure that novice users do not
  count the tails but the etch pits of fission tracks, we will add a
  `quiz' to the tutorial pages. Only users who click on the `correct'
  features in an unlabelled set of images will be allowed to count new
  samples. Section~\ref{sec:crowdsourcing} shows that, despite the
  large scatter of the raw track counts obtained by the citizen
  scientists, their pairwise ratio estimates scatter evenly around the
  expert opinion. This suggests that the collective wisdom of the
  entire group of students is greater than that of the individual
  analysts. When applied to a group that has achieved a good level of
  training and experience, crowd sourcing should improve the precision
  and accuracy of fission track data.
\item Length measurements. The geochron@home archive already
  contains horizontally confined fission track measurements
  \citep{tamer2025}.  However, these results must be generated
  externally (e.g., using Fission Track Studio) and uploaded via a
  \texttt{.json} file.  The virtual microscope curently lacks the
  functionality to generate length data within
  geochron@home. This functionality will be added in a future
  update.
\item Dpar and Dper. Etch pits are currently stored as simple sets of
  x- and y-coordinates.  In reality, etch pits have a finite length
  (`Dpar') and width (`Dper'), which serve as useful indicators for
  the horizontal etch rates along the c-axis and parallel to it
  \citep{donelick1993}. Functions will be added to measure and
  visualise this type of data in geochron@home.
\item Machine learning. Data science is experiencing an artificial
  intelligence revolution that has already started to transform the
  fission track method \citep{nachtergaele2020}.  Convolutional neural
  networks must be trained with example data. geochron@home
  is ideally suited for this task. Section~\ref{sec:crowdsourcing}
  showed how the collective wisdom of multiple fission track analysts
  can label fission track images with probabilities rather than
  counts. This data format is close to the form in which data are
  treated within an AI algorithm.\medskip

  Once trained on historical data, AI algorithms can be used to count
  fission tracks automatically. Following the model of Fission Track
  Studio \citep{gleadow2009, gleadow2019}, machine learning can be
  used to reverse the fission track counting process. Instead of
  asking users to count the fission tracks in a sample, the software
  can ask them to check the results proposed by an AI algorithm, and
  to remove any features that are \emph{not} fission tracks.\medskip

  Regardless of whether fission tracks were counted manually or with a
  machine, the value of the geochron@home archive remains the
  same. It is important to document data so that samples can be
  reanalysed in the future, for example when a new and improved
  generation of machine learning algorithms becomes available.
\end{enumerate}

These four improvements will be made by ourselves pending additional
funding. However, because geochron@home is free and open software, we
invite any interested parties to join the effort and extend or improve
our code.

@@

Another advantage of the geochron@home wokflow is the separation of
image acquisition and image analysis. This separation reduces the
hardware requirements for fission track geochronology. It opens up the
possibility to share resources. State-of-the-art digital microscopes
are expensive. Using geochron@home, a single microscope can serve
multiple users and make fission track analysis more
affordable.\medskip

>>>>

\begingroup
\captionof{table}{Summary table of the crowd sourcing results for the
  different users (rows) and grains (columns).  The last row is the
  ratio of the median count to PV's count for each grain.  The last
  column is the ratio of the pooled count over grains compared with PV
  for the same grains.}
\label{tab:crowdtable}
\renewcommand{\arraystretch}{0.7}
\begin{scriptsize}
  \input{../output/crowdtable.txt}
\end{scriptsize}
\endgroup

In 1906, Sir Francis Galton visited a county fair in which a contest
was held to guess the weight of an ox. 787 people participated in the
event. Galton discovered that the median of all their estimates was
within 0.8\% of the true weight of the ox and more accurate than 90\%
of the individual estimates. Such is the `wisdom of crowds'
\citep{galton1907a}. Similar effects are seen in fission track
geochronology. An interlaboratory comparison study by
\citet{miller1985} showed that the average of several fission track
age estimates is closer to the known age of mineral standards than the
age obtained by individual observers. Inspired by these previous
experiments, we used geochron@home to bring fission track analysis to
a proverbial `county fair' of citizen scientists.\medskip

We here present the results of a type of `crowd-sourcing' experiment
that was carried out at UCL as part of an undergraduate course in
isotope geology. 70 students were asked to create an account on the
geochron@home server by selecting a unique username and
password. After spending a few minutes to complete the compulsory
tutorial (Section~\ref{sec:tutorial}), they were asked to count
fission tracks in randomly assigned image stacks of Mount Dromedary
apatite. Each student was required to analyse at least 15 grains,
using pre-defined ROIs. In a matter of hours, the students amassed a
dataset of 37,331 fission track counts in 34 grains.  This large
dataset was parsed into separate data files --- one for each student
--- which they processed during an assessed programming
exercise.\medskip

@@

However, in spite of the under-counting by the students, the
\emph{ratios} of the track counts between different grains are
generally quite close to PV's ratios, especially when comparing grains
with similar track densities.  This is apparent from
Table~\ref{tab:crowdtable} (looking along columns).  In grains
containing few tracks (left hand side of the table), the median of the
student counts is within 20\% of PV's counts.  However, this gradually
levels off to 50\% at higher track densities (right hand side of the
table). Figure~\ref{fig:23vs25}b shows a scatter plot of pairs of
counts for two grains (23 and 25) made by 36 students who counted both
of them. The two grains were chosen because they have similar track
counts for both PV and the median student. They both have a median of
28 tracks per grain compared with PV's pair of 52 and 50, which is
highlighted in blue. There is a positive correlation between the
pairs, consistent with systematic observer effects (i.e., people who
count a low value in one grain tend to count similarly low in the
other, and vice versa).  Comparing other pairs of grains shows similar
results, with correlations varying between 0.4 and 0.9. Pairs of
grains with similar track densities are more strongly correlated than
those with dissimilar track densities. In all cases we have looked at,
the residual variation, after allowing for systematic differences
between students, is consistent with Poisson variation.\medskip

As mentioned in Section~\ref{sec:architecture}, geochron@home stores
the actual track positions marked by the users. This raw data can be
downloaded as a \texttt{.json} file and inspected in detail.
Figure~\ref{fig:23vs25}b) shows two-dimensional histograms for the x-y
positions of all the track positions generated by the students in
grains~23 and 25. Visual comparison of the histogram with the optical
image confirms that the students unanimously identified the most
obvious semitracks, which contain a clearly visible etch pit and
tail. Shorter and fainter tracks received fewer clicks. Datasets like
this can be used to replace integer counts of fission tracks with
probabilities, reflecting the ambiguity of some fission track
datasets. Figure~\ref{fig:23vs25} also shows that some students
counted the tails of the fission tracks rather than their etch pits,
despite being told the opposite in the tutorial. Fixing this issue
will require some improvements to the tutorial pages
(Section~\ref{sec:outlook}).\medskip

%\begin{figure}[!ht]
{ \centering \includegraphics[width=15cm]{23vs25.pdf}
  \captionof{figure}{ a) The track counts for grain~25 plotted against
    those for grain~23 for 37 students who counted both of those
    grains. b) Optical images in transmitted light of Mount Dromedary
    apatites 23 and 25 in the crowd-sourcing experiment, along with
    two-dimensional histogram of the track locations for the two
    grains, as identified by the citizen scientists.
    \medskip}
  \label{fig:23vs25}
}%\end{figure}

@@

\begin{enumerate}
\item Length measurements. The geochron@home archive already
  contains horizontally confined fission track measurements
  \citep{tamer2025}.  However, these results must be generated
  externally (e.g., using Fission Track Studio) and uploaded via a
  \texttt{.json} file.  The virtual microscope curently lacks the
  functionality to generate length data within
  geochron@home. This functionality will be added in a future
  update.
\item Dpar and Dper. Etch pits are currently stored as simple sets of
  x- and y-coordinates.  In reality, etch pits have a finite length
  (`Dpar') and width (`Dper'), which serve as useful indicators for
  the horizontal etch rates along the c-axis and parallel to it
  \citep{donelick1993}. Functions will be added to measure and
  visualise this type of data in geochron@home.
\item Interactive tutorial pages and embedded quality
  control. Section~\ref{sec:crowdsourcing} shows that the
  crowd-sourcing idea is not yet ready for general use. Two
  improvements will be made to make its results more reliable. First,
  a `quiz' to the tutorial pages to ensure that novice users do not
  count the tails but the etch pits of fission tracks. Only users who
  click on the `correct' features in an unlabelled set of images will
  be allowed to count new samples. Second, future citizen scientists
  will occasionally be served reference images of known track
  density. Their counts for these `test' images will be used to
  identify trustworthy analysts (top of Table~\ref{tab:crowdtable}),
  and flag unreliable ones (bottom of Table~\ref{tab:crowdtable}).
\item Machine learning. Data science is experiencing an artificial
  intelligence revolution that has already started to transform the
  fission track method \citep{nachtergaele2020}.  Convolutional neural
  networks must be trained with example data. geochron@home
  is ideally suited for this task. Section~\ref{sec:crowdsourcing}
  showed how the collective wisdom of multiple fission track analysts
  can label fission track images with probabilities rather than
  counts. This data format is close to the form in which data are
  treated within an AI algorithm.
\item Inverse counting. Once trained on historical data, AI algorithms
  can be used to count fission tracks automatically. Following the
  model of Fission Track Studio \citep{gleadow2009, gleadow2019},
  machine learning can be used to reverse the fission track counting
  process. Instead of asking users to count the fission tracks in a
  sample, the software can ask them to check the results proposed by
  an AI algorithm, and to remove any features that are \emph{not}
  fission tracks.  Regardless of whether fission tracks were counted
  manually or with a machine, the value of the geochron@home archive
  remains the same. It is important to document data so that samples
  can be reanalysed in the future, for example when a new and improved
  generation of machine learning algorithms becomes available.
\end{enumerate}

These improvements will be made by ourselves pending additional
funding. However, because geochron@home is free and open software, we
invite any interested parties to join the effort and extend or improve
our code.

@@

Another advantage of the geochron@home workflow is the separation of
image acquisition and image analysis. This separation reduces the
hardware requirements for fission track geochronology. It opens up the
possibility to share resources. State-of-the-art digital microscopes
are expensive. Using geochron@home, a single microscope can serve
multiple users and make fission track analysis more
affordable.\medskip
